Physics of Tidal Dissipation in Early-Type Stars and White Dwarfs: Hydrodynamical Simulations of Internal Gravity Wave Breaking in Stellar Envelopes

### 2020-04-04 writeups/Su_IGW_break.tex
check in comments, add typo fixes

Tidal dissipation can also play an important role in compact double white dwarf
(WD) binary systems (with orbital periods in the range of minutes to hours).
Such binaries may produce a variety of exotic astrophysical systems and
phenomena, ranging from isolated sdB/sdO stars, R CrB stars, AM CVn binaries,
high-mass neutron stars and magnetars (created by the accretion-induced collapse
of merging WDs), and various optical transients (underluminous supernovae,
Ca-rich fast transients, and type Ia supernovae)
\citep[e.g.][]{livio2018progenitors,toloza2019understanding}. The outcomes of WD
mergers depend on the WD masses and composition, but tidal dissipation can
strongly affect the pre-merger conditions of the WDs and therefore the merger
outcomes. Tidal dessipation may also influence the evoluton of eccentric
WD-massive black hole binaries prior to the eventual tidal disruption of the WD
\citep{vick2017tidal}.

Recent studies have identified nonlinear dissipation of IGWs as the key tidal
process in compact WD binaries \citep{fullerII,fullerIV,tidal_novae,bukart}:
IGWs are tidally excited mainly at the compositon transitions of the WD
envelope; as these waves propagate outwards towards the WD surface, they grow in
amplitude until they break, and transfer both energy and angular momentum from
the binary orbit to the outer envelope of the WD\@. However, these previous
works parameterized the wave breaking process in an ad hoc manner. The details
of dissipation, namely the location and spatial extent of the wave breaking,
affect the observable outcomes: dissipation near the surface of the WD can be
efficiently radiated away and simply brightens the WD, while dissipation deep in
the WD envelope causes an energy buildup that results in energetic flares
\citep{tidal_novae}. An important goal of this paper is to elucidate the details
of the nonlinear IGW breaking process; the result of this ``microphysics'' study
will help determine the thermal evolution and the observational manifestations
of tidally heated binary WDs.

@@

\citet{drazin, klostermeyer, winters1994} describe the onset of wave breaking in
some detail. At intermediate amplitudes, wave breaking occurs via triadic
resonances, transfering energy from the ``parent'' IGW to ``daughter'' waves on
smaller length scales that efficiently damp. The horizontal momentum flux
decreases from $F$ to $0$ over this breaking region. The lost flux is deposited
into a horizontal mean flow
\begin{equation}
    \overline{U}(z, t) \equiv \ev{u_x}_x.\label{eq:mean_flow}
\end{equation}
As the mean flow grows, a \emph{critical layer} may form, as discussed below.

@@

The analytical theory (Section~\ref{ss:lin_analysis}) also predicts that the
horizontal momentmum flux $F(z, t)$ is independent of $z$ between the forcing
zone where the wave is generated and the damping zone where it is dissipated.
The expected horizontal momentum flux carried by the excited IGW in the linear
theory can be computed by simply evaluating Eq.~\eqref{eq:F_def} for
$\bm{u}_{al}$ and is a constant:
\begin{equation}
    F_{al} \equiv \ev{\rho u_{al, x} u_{al, z}}_x.\label{eq:F_al}
\end{equation}
Denoting the momentum flux measured in the simulation by $F(z, t)$
(use Eq.~\eqref{eq:F_def} with velocities taken from the simulation), we expect
$F(z, t) = F_{al}$ between $z_0$ and $z_T$. Fig.~\ref{fig:lin_fluxes} shows
agreement with this prediction.
\begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{plots/lin_fluxes.png}
    \caption{$F/F_{al}$ plotted at select times $t$ (in units of
    $N^{-1}$). As the initial transient dies out, $F / F_{al} \approx 1$ to a
    good approximation above the forcing zone $z > z_0 = 2H$ and below the
    damping zone $z \lesssim z_T = 9.5H$. The horizonal momentum flux excited in
    the forcing zone is transported without loss to the top of the domain, where
    it is dissipated by the damping zone (see Section~\ref{ss:damping}) without
    reflection.}\label{fig:lin_fluxes}
\end{figure}

@@

Nevertheless, the difference in behavior of $\mathrm{Ri}$ and the flux
reflectivity as $\mathrm{Re}$ is varied is in tension with
Eq.~\eqref{eq:crit_coeffs}. This tension is natural: Eq.~\eqref{eq:crit_coeffs}
is derived from a linear theory, while fluid motion within the critical layer is
turbulent, so reflecton at the critical layer cannot be captured by the linear
theory.

@@

In this paper, we have studied the nonlinear behavior of IGWs with $\abs{k_x /
k_z} \sim 1 / (2\pi)$ in a plane-parallel geometry. Tidally excited IGWs in
binary stars have horizontal wave number $k_\perp \sim 1/R$ (where $R$ is the
stellar radius) much smaller than the radial wave number $k_r$. While our
simulations do not satisfy $\abs{k_x/k_z}\ll 1$, the qualitative behavior is likely
to be similar, as the turbulence driving the critical layer dynamics occurs at
scales significantly smaller than either $1/k_x$ or $1/\abs{k_z}$. Simulating
IGWs with $k_x \ll \abs{k_z}$ is more challenging numerically and we defer its
exploration to future work.

@@

The system of equations we wish to simulate consists of
Eqs.~\eqref{eq:nl_incomp},~\eqref{eq:nl_upsilon_u}, and~\eqref{eq:vol_drive}.
% can be explicitly written in
% component form and partial derivatives below:
% \begin{subequations}\label{se:nl_var}
%     \begin{align}
%         \bm{\nabla} \cdot \bm{u} &= 0,\\
%         \pd{\Upsilon}{t} + \p{\bm{u} \cdot \bm{\nabla}} \Upsilon
%             - \frac{u_z}{H} &= 0,\\
%         \pd{u_{x}}{t} + \p{\bm{u} \cdot \bm{\nabla}}u_{x}
%             + \pd{\varpi'}{x} + gH\pd{\Upsilon}{x}
%             + \varpi' \pd{\Upsilon}{x} &= 0,\\
%         \pd{u_z}{t} + \p{\bm{u} \cdot \bm{\nabla}}u_z
%             + \pd{\varpi'}{z} + gH\pd{\Upsilon}{z}
%             + \varpi' \pd{\Upsilon}{z} - \frac{\varpi'}{H} &= 0.
%     \end{align}
% \end{subequations}
The nonlinear terms in the these equations will transfer energy from lower
wavenumbers to higher wavenumbers. Since spectral codes have no numerical
diffusion, explicit diffusion must be added. To ensure the non-ideal
system conserves horizontal momentum exactly, we begin by adding diffusion
terms to the flux-conservative form of the Euler fluid equations
(equivalent to Eqs.~\ref{se:nl_orig}):
\begin{subequations}
    \begin{align}
        \bm{\nabla} \cdot \bm{u} &= 0,\\
        \partial_t \rho + \bm{\nabla} \cdot (\rho \bm{u} - \nu
            \bm{\nabla}(\rho - \overline{\rho})) &= 0,\label{eq:visc_cons_mom}\\
        \partial_t (\rho \bm{u}) + \bm{\nabla} \cdot (\rho \bm{u} \bm{u}
            + \mathrm{diag}(\rho \varpi)
            - \nu \rho \bm{\nabla}\bm{u})
            + \rho g \uv{z} &= 0.
    \end{align}
\end{subequations}
For simplicity, we use the same diffusivity $\nu$ for both the momentum and mass
diffusivities. Although mass diffusivity is not physical, we incclude it for
numerical stability. We choose the mass diffusion term to conserve mass, and not
to affect the background density profile.

>>>>

Tidal dissipation can also play an important role in compact double white dwarf
(WD) binary systems (with orbital periods in the range of minutes to hours).
Such binaries may produce a variety of exotic astrophysical systems and
phenomena, ranging from isolated sdB/sdO stars, R CrB stars, AM CVn binaries,
high-mass neutron stars and magnetars (created by the accretion-induced collapse
of merging WDs), and various optical transients (underluminous supernovae,
Ca-rich fast transients, and type Ia supernovae)
\citep[e.g.][]{livio2018progenitors,toloza2019understanding}. The outcomes of WD
mergers depend on the WD masses and composition, but tidal dissipation can
strongly affect the pre-merger conditions of the WDs and therefore the merger
outcomes. Tidal dissipation may also influence the evolution of eccentric
WD-massive black hole binaries prior to the eventual tidal disruption of the WD
\citep{vick2017tidal}.

Recent studies have identified nonlinear dissipation of IGWs as the key tidal
process in compact WD binaries \citep{fullerII,fullerIV,tidal_novae,bukart}:
IGWs are tidally excited mainly at the composition transitions of the WD
envelope; as these waves propagate outwards towards the WD surface, they grow in
amplitude until they break, and transfer both energy and angular momentum from
the binary orbit to the outer envelope of the WD\@. However, these previous
works parameterized the wave breaking process in an ad hoc manner. The details
of dissipation, namely the location and spatial extent of the wave breaking,
affect the observable outcomes: dissipation near the surface of the WD can be
efficiently radiated away and simply brightens the WD, while dissipation deep in
the WD envelope causes an energy buildup that results in energetic flares
\citep{tidal_novae}. An important goal of this paper is to elucidate the details
of the nonlinear IGW breaking process; the result of this ``microphysics'' study
will help determine the thermal evolution and the observational manifestations
of tidally heated binary WDs.

@@

\citet{drazin, klostermeyer, winters1994} describe the onset of wave breaking in
some detail. At intermediate amplitudes, wave breaking occurs via triadic
resonances, transferring energy from the ``parent'' IGW to ``daughter'' waves on
smaller length scales that efficiently damp. The horizontal momentum flux
decreases from $F$ to $0$ over this breaking region. The lost flux is deposited
into a horizontal mean flow
\begin{equation}
    \overline{U}(z, t) \equiv \ev{u_x}_x.\label{eq:mean_flow}
\end{equation}
As the mean flow grows, a \emph{critical layer} may form, as discussed below.

@@

The analytical theory (Section~\ref{ss:lin_analysis}) also predicts that the
horizontal momentum flux $F(z, t)$ is independent of $z$ between the forcing
zone where the wave is generated and the damping zone where it is dissipated.
The expected horizontal momentum flux carried by the excited IGW in the linear
theory can be computed by simply evaluating Eq.~\eqref{eq:F_def} for
$\bm{u}_{al}$ and is a constant:
\begin{equation}
    F_{al} \equiv \ev{\rho u_{al, x} u_{al, z}}_x.\label{eq:F_al}
\end{equation}
Denoting the momentum flux measured in the simulation by $F(z, t)$
(use Eq.~\eqref{eq:F_def} with velocities taken from the simulation), we expect
$F(z, t) = F_{al}$ between $z_0$ and $z_T$. Fig.~\ref{fig:lin_fluxes} shows
agreement with this prediction.
\begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{plots/lin_fluxes.png}
    \caption{$F/F_{al}$ plotted at select times $t$ (in units of
    $N^{-1}$). As the initial transient dies out, $F / F_{al} \approx 1$ to a
    good approximation above the forcing zone $z > z_0 = 2H$ and below the
    damping zone $z \lesssim z_T = 9.5H$. The horizontal momentum flux excited
    in the forcing zone is transported without loss to the top of the domain,
    where it is dissipated by the damping zone (see Section~\ref{ss:damping})
    without reflection.}\label{fig:lin_fluxes}
\end{figure}

@@

Nevertheless, the difference in behavior of $\mathrm{Ri}$ and the flux
reflectivity as $\mathrm{Re}$ is varied is in tension with
Eq.~\eqref{eq:crit_coeffs}. This tension is natural: Eq.~\eqref{eq:crit_coeffs}
is derived from a linear theory, while fluid motion within the critical layer is
turbulent, so reflection at the critical layer cannot be captured by the linear
theory.

@@

In this paper, we have studied the nonlinear behavior of IGWs with $\abs{k_x /
k_z} \sim 1 / (2\pi)$ in a plane-parallel geometry. Tidally excited IGWs in
binary stars have horizontal wavenumber $k_\perp \sim 1/R$ (where $R$ is the
stellar radius) much smaller than the radial wavenumber $k_r$. While our
simulations do not satisfy $\abs{k_x/k_z}\ll 1$, the qualitative behavior is likely
to be similar, as the turbulence driving the critical layer dynamics occurs at
scales significantly smaller than either $1/k_x$ or $1/\abs{k_z}$. Simulating
IGWs with $k_x \ll \abs{k_z}$ is more challenging numerically and we defer its
exploration to future work.

@@

The system of equations we wish to simulate consists of
Eqs.~\eqref{eq:nl_incomp},~\eqref{eq:nl_upsilon_u}, and~\eqref{eq:vol_drive}.
% can be explicitly written in
% component form and partial derivatives below:
% \begin{subequations}\label{se:nl_var}
%     \begin{align}
%         \bm{\nabla} \cdot \bm{u} &= 0,\\
%         \pd{\Upsilon}{t} + \p{\bm{u} \cdot \bm{\nabla}} \Upsilon
%             - \frac{u_z}{H} &= 0,\\
%         \pd{u_{x}}{t} + \p{\bm{u} \cdot \bm{\nabla}}u_{x}
%             + \pd{\varpi'}{x} + gH\pd{\Upsilon}{x}
%             + \varpi' \pd{\Upsilon}{x} &= 0,\\
%         \pd{u_z}{t} + \p{\bm{u} \cdot \bm{\nabla}}u_z
%             + \pd{\varpi'}{z} + gH\pd{\Upsilon}{z}
%             + \varpi' \pd{\Upsilon}{z} - \frac{\varpi'}{H} &= 0.
%     \end{align}
% \end{subequations}
The nonlinear terms in the these equations will transfer energy from lower
wavenumbers to higher wavenumbers. Since spectral codes have no numerical
diffusion, explicit diffusion must be added. To ensure the non-ideal
system conserves horizontal momentum exactly, we begin by adding diffusion
terms to the flux-conservative form of the Euler fluid equations
(equivalent to Eqs.~\ref{se:nl_orig}):
\begin{subequations}
    \begin{align}
        \bm{\nabla} \cdot \bm{u} &= 0,\\
        \partial_t \rho + \bm{\nabla} \cdot (\rho \bm{u} - \nu
            \bm{\nabla}(\rho - \overline{\rho})) &= 0,\label{eq:visc_cons_mom}\\
        \partial_t (\rho \bm{u}) + \bm{\nabla} \cdot (\rho \bm{u} \bm{u}
            + \mathrm{diag}(\rho \varpi)
            - \nu \rho \bm{\nabla}\bm{u})
            + \rho g \uv{z} &= 0.
    \end{align}
\end{subequations}
For simplicity, we use the same diffusivity $\nu$ for both the momentum and mass
diffusivities. Although mass diffusivity is not physical, we include it for
numerical stability. We choose the mass diffusion term to conserve mass, and not
to affect the background density profile.
